Implementing Equational Constraints\\ in a Functional Language

### 2011-06-30 papers/wlp2011/paper.tex
Benchmark chapter extended

A well-known disadvantage of equational constraints
is the fact that \ccode{=:=} is interpreted as strict equality.
Thus, if one uses equational constraints to express
requirements on arguments, the resulting operations might be too strict.
For instance, the equational constraint in the condition
defining \code{last} (see Section~\ref{ex:last})
requires that \code{ys++[e]} as well as \code{xs}
must be reducible to identical values so that the input list \code{xs}
is completely evaluated.
Hence, if \code{failed} denotes an operation whose evaluation fails,
the evaluation of \code{last [failed,True]} has no result.
On the other hand, the evaluation of \code{last' [failed,True]} yields
the value \code{True}, i.e., the definition of \code{last'} is less strict.
This is due to the fact that a functional pattern like
\code{(xs++[e])} abbreviates all values to which it can be evaluated
(by narrowing), like \code{[e]}, \code{[x1,e]}, \code{[x1,x2,e]} etc.
Conceptually, the rule defining \code{last'}
abbreviates the following (infinite) set of rules:
\begin{curry}
  last' [e] = e
  last' [x1,e] = e
  last' [x1,x2,e] = e
  $\ldots$
\end{curry}
As a further example,
consider an operation that returns the first duplicated element in
a list.
Using equational constraint, we can define it as follows:
\begin{curry}
  fstDup xs | xs =:= ys++[e]++zs & elem e ys =:= True & nub ys =:= ys
            = e    where ys,zs,e free
\end{curry}
The first equational constraint is used to split the input list \code{xs}
into three sublists.
The last constraint ensures that the first sublist
\code{ys} does not contain duplicated elements
(the library operation removes all duplicates from a list)
and the second constraint ensures that the first element after \code{ys}
occurs in \code{ys}.
Although this implementation is concise, it cannot be applied
to infinite lists due to the strict interpretation of \ccode{=:=}.
This is not the case if we define this operation by a functional pattern:
\begin{curry}
  fstDup' (ys++[e]++zs) | elem e ys =:= True & nub ys =:= ys
                        = e
\end{curry}
%
Functional patterns are quite useful in situations
where one wants to select parts of some structure
at arbitrary positions. For instance,
functional patterns have been applied to support
the high-level programming of querying or transforming
XML data \cite{Hanus11ICLP}.

@@

This discussion shows the requirements on an implementation
of functional patterns. The general structure
is quite similar to equational constraints
with the exception that variables could be also bound
to unevaluated expressions. If such variables are later
accessed, they are evaluated.
This can be achieved by adding a further alternative
to the type of decisions:
\begin{haskell}
  data Decision = $\ldots$ | LazyBind [Constraint]
\end{haskell}
%
The implementation of the lazy unification operation \ccode{=:<=}
is almost identical to the strict unification operation \ccode{=:=}
as shown in Section~\ref{sec:Unification}.
The only change is in the rules where a free variable occurs
in the left argument. All these rules are replaced by the single
rule
\begin{haskell}
  Choice (FreeID i) _ _ =:<= x
    = Guard [i :=: LazyBind (lazyBind i x)] Success
\end{haskell}
where the auxiliary operation \code{lazyBind}
implements the demand-driven evaluation of the right argument \code{x}:
\begin{haskell}
  lazyBind :: ID -> a -> [Constraint]
  lazyBind i True  = [i :=: ChooseLeft ]
  lazyBind i False = [i :=: ChooseRight]
\end{haskell}
This has the effect that the argument \code{x} is stored
without evaluation (due to the lazy evaluation strategy of the
target language Haskell) as a binding constraint.
However, it is evaluated by \code{lazyBind} when its binding
is required by another part of the computation.
Similarly to equational constraints,
lazy bindings are processed by a solver when values are extracted.
In particular, if a variable has more than one lazy binding constraint
(which is possible if a functional pattern evaluates to a non-linear term),
the corresponding expressions are evaluated and unified
according to the semantics of functional patterns
\cite{AntoyHanus05LOPSTR}.

In order to demonstrate the operational behavior of our implementation,
we sketch the evaluation of the lazy unification constraint
\code{xs++[e]=:<=[failed,True]} that occurs when the expression
\code{last' [failed,True]} is evaluated (here we omit failed branches
and some other details; note that logic variables are replaced
by generators):
\begin{haskell}
  aBList 2 ++ [aBool 3] =:<= [failed,True]
  $\leadsto$ [aBool 4, aBool 3] =:<= [failed, True]
  $\leadsto$ aBool 4 =:<= failed & aBool 3 =:<= True & [] =:<= []
  $\leadsto$ Guard [ 4 :=: LazyBind (lazyBind 4 failed)
           , 3 :=: LazyBind (lazyBind 3 True)] Success
\end{haskell}
If the binding of identifier \code{3} is later required,
\code{(lazyBind 3 True)} is evaluated to \code{[3 :=: ChooseLeft]}
which corresponds to the value \code{True} of the generator \code{aBool}.
Note that the variable with identifier \code{4} does not occur
anywhere else so that the binding \code{(lazyBind 4 failed)}
will never be evaluated, as intended.

@@

Benchmarks for Last with ==, =:= and func. patterns

also short comparison of =:= with PAKCS and MCC

>>>>

A well-known disadvantage of equational constraints
is the fact that \ccode{=:=} is interpreted as strict equality.
Thus, if one uses equational constraints to express
requirements on arguments, the resulting operations might be too strict.
For instance, the equational constraint in the condition
defining \code{last} (see Section~\ref{ex:last})
requires that \code{ys++[e]} as well as \code{xs}
must be reducible to identical values so that the input list \code{xs}
is completely evaluated.
Hence, if \code{failed} denotes an operation whose evaluation fails,
the evaluation of \code{last [failed,2]} has no result.
On the other hand, the evaluation of \code{last' [failed,2]} yields
the value \code{2}, i.e., the definition of \code{last'} is less strict.
This is due to the fact that a functional pattern like
\code{(xs++[e])} abbreviates all values to which it can be evaluated
(by narrowing), like \code{[e]}, \code{[x1,e]}, \code{[x1,x2,e]} etc.
Conceptually, the rule defining \code{last'}
abbreviates the following (infinite) set of rules:
\begin{curry}
  last' [e] = e
  last' [x1,e] = e
  last' [x1,x2,e] = e
  $\ldots$
\end{curry}
As a further example,
consider an operation that returns the first duplicated element in
a list.
Using equational constraint, we can define it as follows:
\begin{curry}
  fstDup xs | xs =:= ys++[e]++zs & elem e ys =:= True & nub ys =:= ys
            = e    where ys,zs,e free
\end{curry}
The first equational constraint is used to split the input list \code{xs}
into three sublists.
The last constraint ensures that the first sublist
\code{ys} does not contain duplicated elements
(the library operation removes all duplicates from a list)
and the second constraint ensures that the first element after \code{ys}
occurs in \code{ys}.
Although this implementation is concise, it cannot be applied
to infinite lists due to the strict interpretation of \ccode{=:=}.
This is not the case if we define this operation by a functional pattern:
\begin{curry}
  fstDup' (ys++[e]++zs) | elem e ys =:= True & nub ys =:= ys
                        = e
\end{curry}
%
Functional patterns are quite useful in situations
where one wants to select parts of some structure
at arbitrary positions. For instance,
functional patterns have been applied to support
the high-level programming of querying or transforming
XML data \cite{Hanus11ICLP}.

@@

This discussion shows the requirements on an implementation
of functional patterns. The general structure
is quite similar to equational constraints
with the exception that variables could be also bound
to unevaluated expressions. If such variables are later
accessed, they are evaluated.
This can be achieved by adding a further alternative
to the type of decisions:
\begin{haskell}
data Decision = $\ldots$ | LazyBind [Constraint]
\end{haskell}
%
The implementation of the lazy unification \ccode{=:<=}
is almost identical to the strict unification \ccode{=:=}

@@

Unlike the Curry implementations PAKCS and MCC our approach does
not manipulate a global state to unify a logic variable. Nevertheless,
our benchmarks show that it can compete or even outperform these 
implementations.
The programs used for the benchmarks are primarily the same as in \cite{AntoyHanus05LOPSTR}.

\begin{tabular}{l l}
\bf{last}& Computes the last element of a list using strict equality\\
\bf{simplify}& Simplification of symbolic arithmetic expressions\\
\bf{varInExp}& Nondeterministically returns a variable occuring
                in a symbolic arithmetic expression \\
\bf{half}& Computes the half of a peano number using logic variables\\
\end{tabular}

The table below shows a comparison of the curry systems where
strict equality (\code{=:=}) is used for unification:
\footnote{\ccode{inc x n} is a naive addition that adds 1 \code{n} times
          to its argument x }

\begin{tabular}{|l|c|c|c|}
\hline
Expression: & KiCS2 & PAKCS & MCC \\
\code{last (map (inc 0) [1..100000])} & &  & \\
\code{simplify}                       &&&\\
\code{varsInExp} & &  &\\
\code{toPeano 5000 == halfS (toPeano 10000)} & &  &\\
\hline
\end{tabular}

All of our examples can also be defined with function patterns, i.e.,
the lazy equality operator \code{=:<=} is used. The sole other
Curry system featuring function patterns is PAKCS:

\begin{tabular}{|l|c|c|c|}
\hline
Expression: & KiCS2 & PAKCS \\
\code{last (map (inc 0) [1..100000])} & &  & \\
\code{simplify}                       &&&\\
\code{varsInExp} & &  &\\
\code{toPeano 5000 == halfS (toPeano 10000)} & &  &\\
\hline
\end{tabular}

In Section \ref{sec:Unification} we stated that equational
constraits can be implemented using ordinary pattern matching.
We also mentioned that such an implementation would lead to
a large search space that has to be traversed to extract
the results of a computation. To show the benificial effects
of our equality constraint implementations we compare them to
the results using the equality operator \code{==} that
produces such a search space:

\begin{tabular}{|l|c|c|c|}
\hline
Expression: &\code{==} & \code{=:=} & \code{=:<=}\\
\hline 
\code{last (map (inc 0) [1..100000])} & stack overflow & 0.45 & 0.13\\
\code{simplify}                       &10.34&10.33&7.27\\
\code{varsInExp} & 0.49& 84.92 & 0.09\\
\code{toPeano 5000 == halfS (toPeano 10000)} & 23.3 & 1.81 & no Fun Pat\\
\hline
\end{tabular}
